\textbf{COMP0083 Convex Optimisation Assignment}

### 2023-01-07 submission.tex
add report edits

The function for (a):

\[\max\{ax+b, x^4-5, e^{x^2}\}\]

@@

For the function:

\[f(x) = \begin{cases}
      -x \tab \text{if } x \in ]-1, 0] \\
      x^2 \tab \text{if } x \geq 0 \\
   \end{cases}\]

the sub-differential is:

\[\partial f(x) = \begin{cases}
      -1 \tab \text{if } x \in ]-1, 0[ \\
      \left[0, 1\right] \tab \text{if } x = 0  \\
      2x \tab \text{if } x > 0 \\
   \end{cases}\]

@@

\subsection*{1.3}
For a function:
\[f(x) = \langle Ax, x \rangle + \langle x, b\rangle + c\]

where A is a square matrix not necessarily symmetric, the gradient is (a):

\[\nabla f(x) = A^*x + Ax + b\]

\subsection*{1.4}

The Fenchel conjugate of $f(x) = g(2x)$ is (a):

\[f^*(u) = g^*(u/2)\]

\subsection*{1.5}

The solution to the dual problem is (c):

\[\bar{u} = (\textbf{K}+\lambda n \textbf{Id})^{-1} y\]

\section*{Part 2: Theory on convex analysis and optimization}
\subsection*{2.1}

@@

Given :
\[ f(x) = \begin{cases}
      +\infty \tab \text{if } x \leq 0 \\
      -\log x \tab \text{if } x > 0 \\
   \end{cases}\]

The Fenchel conjugate is defined:

@@

To find the supremum, we can take the partial derivative with respect to $x$, set to zero, and solve for x:

@@

Thus, the supremum above is solved when $x=\frac{-1}{u}$:

@@

\[f^*(u) = - (1 + \log u)\]

@@

The Fenchel conjugate is defined:

@@

We can compute the partial derivative:

@@

Thus, the supremum is solved when $x=\frac{u}{2}$:

@@

We can see the Fenchel conjugate is:

@@

@@

for all $x1, x2 \in \mathcal{X}$ and $\lambda_1, \lambda_2 \in \mathbb{R}_+$ with $\lambda_1 + \lambda_2 = 1$. In this base case $n=2$.

Our inductive step will prove that the inequality continues to hold for $n+1$:

@@

We can insert the term $(1-\lambda_{n+1})$:

@@

Rewriting this,

@@

We know that the first term on the right hand side:

@@

Thus can upper bound our previous inequality:

\[f\left(\sum_{i=1}^{n+1} \lambda_i x_i\right)  \leq \sum_{i=1}^n \frac{\lambda_i}{{1-\lambda_{n+1}}}  f \left(x_i\right) + \lambda_{n+1} f\left( x_{n+1}\right) =\sum_{i=1}^{n+1} \frac{\lambda_i}{{1-\lambda_{n+1}}} f \left(x_i\right) \]

proving our inductive step and Jensen's inequality as required. \square

@@

\[\text{f is convex} \leftrightarrow \langle \nabla f(x) - \nabla f(y), x-y \rangle \geq 0\]

@@

\begin{gather*}
    \frac{1}{y} > \frac{1}{x}\\
    x-y>0\\
    \frac{1}{y} - \frac{1}{x} > 0\\
\end{gather*}

@@

If $x<y$ we have the following:
\begin{gather*}
    \frac{1}{y} < \frac{1}{x}\\
    x-y<0\\
    \frac{1}{y} - \frac{1}{x} < 0\\
\end{gather*}

@@

If $x=y$ we have the following:
\begin{gather*}
    \frac{1}{y} = \frac{1}{x}\\
    x-y=0\\
    \frac{1}{y} - \frac{1}{x} = 0\\
\end{gather*}

@@

Thus, $\forall x, y \in dom f, \langle \nabla f(x) - \nabla f(y), x-y \rangle \geq 0$ so $f(x) = -\log(x)$ is convex. \square

@@

\[\frac{1}{n}\sum_{i=1}^n x_i \geq \sqrt[n]{x_1 \cdots x_n}\] \square
\subsection*{2.3}

@@

where $\sum_{i=1}^{m} \lambda_i = 1$. Knowing that $f$ is a convex function on $C$, we can write Jensen's inequality for :

@@

where $\sum_{i=1}^{m} \lambda_i = 1$.

@@

\[\sum_{i=1}^{m} \lambda_i f(a_i) \leq \max_{\{a_i\}_{i=1}^m} f(a_i)\]

@@

\[f\left(\sum_{i=1}^{m} \lambda_i a_i \right) \leq \max_{\{a_i\}_{i=1}^m} f(a_i)\]

So the maximum of the convex function $f$ on $C$ is attained at one of the vertices $a_1, ..., a_m$.
\square

@@

\[\nabla^2 f(x, y) = \begin{bmatrix}
                 2 & 4\\
                 4 & 8
         \end{bmatrix}
\]

we see that the Hessian is positive semi-definite and so the function is jointly convex.

\subsection*{2.5}
  The conditions for the existence of minimizers is that $f$ is closed and coercive.
  The conditions for the uniqueness of minimizers is that $f$ is strictly convex.
Thus for the existence and uniqueness of minimizers for a convex function $f$, we require that $f$ is closed, coercive, and strictly convex.

@@

where $f(x) = \frac{1}{2}\|\textbf{x}\|^2$ and $g(\textbf{Ax}) =  i_{[0, 1]}\left(\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty}\right)$.

@@

\[\min_{\textbf{x}}  \frac{1}{2}\|\textbf{x}\|^2 + i_{[0, 1]}\left(\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty}\right)\]

@@

The Fenchel conjugates are:
\[g^*(\textbf{Ax}) = i^*_{[0, 1]}\left(\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty}\right) = \frac{1}{\epsilon} \| \textbf{A}\textbf{x} - \textbf{b}\|_1\]

@@

Moreover, for $g(\textbf{u}) =  i_{[0, 1]}\left(\frac{1}{\epsilon}\|\textbf{u} - \textbf{b}\|_{\infty}\right)$:
\[\textbf{0}_n \in dom(g)\]

if $\|\textbf{b}\|_{\infty} \leq \epsilon$. Therefore:
\[\textbf{0}_n \in int \big(dom(g)\big)\]

@@

where $\partial f^*$ and $\partial g^*$ are the subgradients of $f^*$ and $g^*$ respectively.

@@

and so
\[\left(\partial\| \textbf{u} - \textbf{b}\|_1\right)_i = \begin{cases}
      \frac{1}{\epsilon}, &  \text{if } | u_i - b_i| > 0 \\
      -\frac{1}{\epsilon}, &  \text{if } | u_i - b_i| < 0 \\
      [-\frac{1}{\epsilon}, \frac{1}{\epsilon}], &  \text{if } | u_i - b| = 0 \\
   \end{cases} \\
\]

where $\left(\partial\| \textbf{u} - \textbf{b}\|_1\right)_i$ is the $i^{th}$ element of $\partial\| \textbf{u} - \textbf{b}\|_1 \in \mathbb{R}^{n \times 1}$

@@

\subsection*{4.4}
Comparing the two algorithms, we can see that both are able to converge to $\alpha$ vectors producing decision boundaries that are relatively successful at separating the majority of the training points for the two classes. However, the FISTA decision boundary misses a couple points for both classes with a decision boundary that is rather close to boundary points of each class. Thus, the quality of the solution is not as nice as the RCPGA where there is ample buffer surrounding the decision boundary for both classes. This would seem that RCPGA can produce better results, however considering the loss function plots for both algorithms, we can see that this comes at a tradeoff of computational resources/speed. FISTA is able to converge in around 50 iterations whereas RCPGA required on the order of $3\times 10^{5}$ iterations. Although each FISTA step involves two steps (to calculate $\alpha_{k+1}$ and $\nu_{k+1}$), this is still a much faster algorithm than RCGPA. Thus, depending on the practical use case, if there is a requirement for algorithm speed, then FISTA may be more appropriate. On the other hand, if we have more resources available, then RCGPA may be more appropriate as it seems to produce more reasonable/accurate results.

>>>>

The function for (a):
\[\max\{ax+b, x^4-5, e^{x^2}\}\]

@@

For the function:
\[f(x) = \begin{cases}
      -x, & \text{if } x \in ]-1, 0] \\
      x^2, & \text{if } x \geq 0 \\
   \end{cases}\]

the sub-differential is:
\[\partial f(x) = \begin{cases}
      -1, & \text{if } x \in ]-1, 0[ \\
      \left[0, 1\right], & \text{if } x = 0  \\
      2x, & \text{if } x > 0 \\
   \end{cases}\]

@@

\subsection*{1.3}
For a function $f(x) = \langle Ax, x \rangle + \langle x, b\rangle + c$ where A is a square matrix not necessarily symmetric, the gradient is (a):
\[\nabla f(x) = A^*x + Ax + b\]
\subsection*{1.4}
The Fenchel conjugate of $f(x) = g(2x)$ is (a):
\[f^*(u) = g^*(u/2)\]

\subsection*{1.5}
The solution to the dual problem is (c):
\[\bar{u} = (\textbf{K}+\lambda n \textbf{Id})^{-1} y\]

\newpage
\section*{Part 2: Theory on convex analysis and optimization}
\subsection*{2.1}

The Fenchel conjugate is defined as:

\[f^*(u) = \sup_{x\in \mathcal{X}} \{ \langle x, u\rangle - f(x) \}\]

@@

Given :
\[ f(x) = \begin{cases}
      +\infty, & \text{if } x \leq 0 \\
      -\log x, & \text{if } x > 0 \\
   \end{cases}\]

Substituting into the Fenchel conjugate definition:

@@

To find the supremum, we can take the partial derivative with respect to $x$, set to zero, and solve for $x$:

@@

Thus, the supremum above is acheived when $x=\frac{-1}{u}$:

@@

\[ f(x) = \begin{cases}
      - (1 + \log (-u), & \text{if } u < 0 \\
      +\infty, & \text{if } u \geq 0 \\
   \end{cases}\]

@@

Substituting into the Fenchel conjugate definition:

@@

We can compute the partial derivative with respect to $x$, set to zero, and solve for $x$:

@@

Thus, the supremum is achieved when $x=\frac{u}{2}$:

@@

We can see that the Fenchel conjugate is:

@@

is the ReLu function.

@@

for all $x_1, x_2 \in \mathcal{X}$ and $\lambda_1, \lambda_2 \in \mathbb{R}_+$ with $\lambda_1 + \lambda_2 = 1$. In this base case $n=2$. Our inductive step will prove that the inequality continues to hold for $n+1$. We begin with:

@@

where $\sum_{i=1}^{n+1} \lambda_i = 1$. We can insert the term $(1-\lambda_{n+1})$:

@@

Substituting the definition for $\bar{x}$:

@@

We know that from the known inequality from the $n$ case:

@@

Thus we can upper bound our previous inequality:

\[f\left(\sum_{i=1}^{n+1} \lambda_i x_i\right)  \leq (1-\lambda_{n+1})\sum_{i=1}^n \frac{\lambda_i}{{1-\lambda_{n+1}}}  f \left(x_i\right) + \lambda_{n+1} f\left( x_{n+1}\right) =\sum_{i=1}^{n+1} \lambda_i f \left(x_i\right) \]

proving our inductive step for $n+1$ and Jensen's inequality as required. \square

@@

\[\text{A function } f \text{ is convex} \leftrightarrow \langle \nabla f(x) - \nabla f(y), x-y \rangle \geq 0\]

@@

\begin{gather*}
    x-y>0\\
    \frac{1}{y} - \frac{1}{x} > 0\\
\end{gather*}

@@

If $x<y$ we have the following:
\begin{gather*}
    x-y<0\\
    \frac{1}{y} - \frac{1}{x} < 0\\
\end{gather*}

@@

If $x=y$ we have the following:
\begin{gather*}
    x-y=0\\
    \frac{1}{y} - \frac{1}{x} = 0\\
\end{gather*}

@@

Thus, $\forall x, y \in dom f, \langle \nabla f(x) - \nabla f(y), x-y \rangle \geq 0$ so by the characterization for differentiable functions, $f(x) = -\log(x)$ is convex. \square

@@

\[\frac{1}{n}\sum_{i=1}^n x_i \geq \sqrt[n]{x_1 \cdots x_n}\]

for all $x_1, ..., x_n \in \mathbb{R}^+$ \square

\subsection*{2.3}

@@

where $\sum_{i=1}^{m} \lambda_i = 1$. Knowing that $f$ is a convex function on $C$, we can write Jensen's inequality:

@@

@@

\[\sum_{i=1}^{m} \lambda_i f(a_i) \leq \max_{i} f(a_i)\]

@@

\[f\left(\sum_{i=1}^{m} \lambda_i a_i \right) \leq \max_{i} f(a_i)\]

So all function evaluations of $f(x)$ for $x \in C$ are less than or equal to the maximum of the function evaluations at the verticies $a_1, ..., a_m$ and so the maximum of the convex function $f$ on $C$ is attained at one of the vertices $a_1, ..., a_m$.
\square

@@

\[\nabla^2 f(x, y) = \begin{bmatrix}
                 2 & -4\\
                 -4 & 8
         \end{bmatrix}
\]

we have our Hessian matrix for which the determinant is $(2)(8) - (-4)^2 \geq 0$ so we see that the Hessian is positive semi-definite and so the function is jointly convex.

\subsection*{2.5}
  The conditions for the existence of minimizers is that $f$ is closed and coercive.
  The conditions for the uniqueness of minimizers is that $f$ is strictly convex.
Thus the minimal sufficient conditions for the existence and uniqueness of minimizers for a convex function $f$, are that $f$ is closed, coercive, and strictly convex.

@@

where $f(\textbf{x}) = \frac{1}{2}\|\textbf{x}\|^2$ and $g(\textbf{Ax}) =  i_{\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty} \leq 1}\left(\textbf{Ax}\right)$, the indicator function of the unit ball for $\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty}$.

@@

\[\min_{\textbf{x}}  \frac{1}{2}\|\textbf{x}\|^2 + i_{\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty} \leq 1}\left(\textbf{Ax}\right)\]

@@

For the Fenchel conjugate of $g(\textbf{Ax})$, we first recognize that the dual of the indicator function is the support function under $\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty}$ which is the definition for the dual norm $\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{1}$.

Thus:

\[g^*(\textbf{Ax}) = i^*_{\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty} \leq 1}\left(\textbf{Ax}\right) = \frac{1}{\epsilon} \| \textbf{A}\textbf{x} - \textbf{b}\|_1\].

The Fenchel conjugate of $f(\textbf{x})$:

@@

Moreover, for $g(\textbf{u}) =  i_{\frac{1}{\epsilon}\|\textbf{u} - \textbf{b}\|_{\infty} \leq 1}\left(\textbf{u}\right)$:
\[\textbf{0}_n \in dom(g)\]

if $\|\textbf{b}\|_{\infty} \leq \epsilon$ and,
\[\textbf{0}_n \in int \big(dom(g)\big)\]

@@

and equivalently,

\[-\textbf{A}^*\textbf{u} \in \partial f(\textbf{x}) \text{ and } \textbf{u} \in \partial g(\textbf{Ax})\]

we will use the formulation involving $\partial f^*$ and $\partial g^*$, the subgradients of $f^*$ and $g^*$ respectively.

@@

and so
\[\left(\partial \frac{1}{\epsilon}\| \textbf{u} - \textbf{b}\|_1\right)_i = \begin{cases}
      \frac{1}{\epsilon}, &  \text{if } | u_i - b_i| > 0 \\
      -\frac{1}{\epsilon}, &  \text{if } | u_i - b_i| < 0 \\
      [-\frac{1}{\epsilon}, \frac{1}{\epsilon}], &  \text{if } | u_i - b| = 0 \\
   \end{cases} \\
\]

where $\left(\partial\| \textbf{u} - \textbf{b}\|_1\right)_i$ is the $i^{th}$ element of $\partial\| \textbf{u} - \textbf{b}\|_1 \in \mathbb{R}^{n \times 1}$, for $i=1, ..., n$.

@@

\subsection*{4.4}
Comparing the two algorithms, we can see that both are able to converge to $\alpha$ vectors producing decision boundaries that are relatively successful at separating the majority of the training points for the two classes. However, the FISTA decision boundary misses a couple points for both classes with a decision boundary that is rather close to boundary points of each class. Thus, the quality of the solution is not as nice as the RCPGA where there are ample buffer zones surrounding the decision boundary for both classes. This would seem that RCPGA can produce better results, however considering the loss function plots for both algorithms, we can see that this comes at a tradeoff of computational resources/speed. FISTA is able to converge in around 50 iterations whereas RCPGA required on the order of $3\times 10^{5}$ iterations. Although each FISTA step involves two steps (to calculate $\alpha_{k+1}$ and $\nu_{k+1}$), this is still a much faster algorithm than RCGPA. Thus, depending on the practical use case, if there is a requirement for algorithm speed, then FISTA may be more appropriate. On the other hand, if we have more resources available, then RCGPA may be more appropriate as it seems to produce more reasonable/accurate results.
